---
title: "STAT 331 Project"
author: "Emily Stokes, Sophia Nicotera, Ben Cox, Wendy Tan"
format: 
  html:
    code-tools: true
    toc: true
    embed-resources: true
    html-table-processing: none
    self-contained: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
code-fold: true
---

## Project Proposal

For our project, we plan to explore the statistical relationship between education and income across countries. Recent trends in the United States suggest a growing negative attitude toward education, with 49% of Americans believing that a college degree is less important now than in the past (Pew Research). On the contrary to this belief, our hypothesized relationship is that a higher percentage in a bachelor's degree in a country will lead to higher daily income, as the citizens are more informed and equipped with field-related knowledge. To investigate this, we selected two datasets from GapMinder: (1) Mean daily household income in 2017 constant international dollars (Note: GapMinder includes a "forecast" in this dataset with predicted values up to the year 2100) and (2) Percentage of the population with bachelor's equivalent attainment

```{r setup}
library(readr)
library(tidyverse)
library(kableExtra)
library(gganimate)
library(gifski)
library(patchwork)
library(broom)
```

Data Preparation:

We first load and clean the data to ensure consistency and eliminate forecasted values (2024-2100).

```{r}
daily_income <- read_csv("mincpcap_cppp.csv")
ba_equiv <- read_csv("se_ter_cuat_ba_zs.csv")
```

Combine all information into one dataset:

-   Each dataset should be turned into a "long format" so that there is one column for country, one for year, and one for the variable of interest so we can join all data based on the country and Year combination and view the daily income and BA equivalent percentage values.

-   Since the last year where real data was recorded was 2023, we will not include years where any value is "predicted" (2024 to 2100).

```{r}
# create a long version for each dataset
di_long <- daily_income |> 
  pivot_longer(cols = '1800':'2100',
               names_to = "Year",
               values_to = "DailyIncome")
ba_long <- ba_equiv |> 
  pivot_longer(cols = "1970":"2023",
               names_to = "Year",
               values_to = "Bachelors_pct")

# merge all the data (including missing values) from the three datasets,
# but only keep information from years where real data was recorded (no later 
# than 2023.)
gapminder <- di_long |> 
  full_join(ba_long,
            by = c("country", "Year")) |> 
  filter(Year < 2024)

gapminder <- gapminder |> mutate(Year = as.numeric(Year))
gapminder <- gapminder |> drop_na(Bachelors_pct, DailyIncome)


```

## Data Visualization

A scatterplot of each data point in the dataset (each Country + Year) with a best fit line shows a positive relationship between the percentage of the population who have completed a Bachelor's equivalent education and the average daily household income. However, the correlation does not appear to be very strong because there is a lot of variation in the points.

```{r}
x_min <- min(gapminder$Bachelors_pct, na.rm = TRUE)
x_max <- max(gapminder$Bachelors_pct, na.rm = TRUE)
y_min <- min(gapminder$DailyIncome, na.rm = TRUE)
y_max <- max(gapminder$DailyIncome, na.rm = TRUE)

BA_plot <- gapminder |>
  ggplot(mapping = aes(x = Bachelors_pct, y = DailyIncome)) + 
  geom_point(alpha = 0.6, size = 2) + 
  geom_smooth(method = "lm", color = "blue", se = FALSE) +  
  scale_x_continuous(limits = c(x_min, x_max)) +  
  scale_y_continuous(limits = c(y_min, y_max)) +  
  labs(x = "Bachelor's Degree Attainment (Percent of Population)",
       y = "Daily Income (PPP Dollars)",
       title = "Daily Income and Population Bachelor's Attainment",
       subtitle = "Year: {frame_time}") +  
  theme_minimal() 

BA_plot

```

When we look at the relationship between the population percentage with Bachelor's equivalent attainment (explanatory variable) and daily household income (response variable), we notice that there is not a lot of data before the year 2000, but after the year 2000 there are enough data to form an obvious pattern. From 2000-2008, there are still only a small amount of data points. They form, linear patterns but there is a lot of variation and from year to year the rate of change switched from positive to negative and back. This is most likely explained by the fact that there are not enough datapoints to see a real pattern. Around 2010 and on, there are many points, and the pattern is still linear with smaller deviation from the regression line, but the rate of change between the two variables is constantly positive even though it varies with each year. However, from 2013 on, the rate of change is the most stable, and the line only varies by a small amount each time. We can conclude that over the years since 2000, the relationship between population bachelor's attainment and daily household income has been positive.

```{r}
animate_plot <- BA_plot + 
  transition_time(Year) + 
  ease_aes("linear") +
  labs(subtitle = "Year: {round(frame_time)}")

animated_plot <- animate(animate_plot, renderer = gifski_renderer(), fps = 7, width = 1000, height = 700, units = "px") 

anim_save("animation.gif", animated_plot)

animated_plot

```


## Linear Regression
We fit a simple linear regression model using the average bachelor's attainment percentage as the explanatory variable and the average daily household income as the response variable for each country (averaging over all existing years in the dataset since there are some missing values). The unit of observation in our regression model is each country, averaged over all available years (2000 and beyond). Only data from 2000 and later was included to ensure consistency and sufficient sample size. Based on the linear regression between a country's average Bachelor's attainment percentage (explanatory) and the country's average daily household income (response), we can model the relationship with:

$$
\hat{Income} = 1.9434 + 1.4878(Bachelor's \ Attainment \ \%) 
$$

Where

-   $\hat{Y}$ represents the predicted average income for a country

-   $\beta_0 = 1.9434$ is the y-intercept for average daily income for a country.

-   $\beta_1 = 1.4878$ represents the average change in mean daily household income (in PPP dollars) for a country for every 1% increase in the average of the country's population percentage with a bachelor's equivalent attainment.

```{r}
country_averages <- gapminder |> 
  group_by(country) |> 
  summarise(bachelors = mean(Bachelors_pct, na.rm = TRUE),
            income = mean(DailyIncome, na.rm = TRUE))
  
bachelors_lm <- lm(income ~ bachelors, 
                 data = country_averages)

broom::tidy(bachelors_lm)|>
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 4)))  |> 
  knitr::kable()|>
  kable_styling(bootstrap_options = c("striped", "hover"))
  
```

## Model Fit
To assess model fit, we calculate the variance in total response values, fitted values, and residuals.

```{r}
bachelor_resid = broom::augment(bachelors_lm, conf.level = 0.95)

bachelor_resid |> 
  summarize(sst = var(income), 
            ssm = var(.fitted), 
            sse = var(.resid)) |> 
  mutate(rsqr = ssm/sst) |> 
  kable(col.names = c("Total Variance", "Fitted Variance", 
                      "Residual Variance", "R^2"), 
        caption = "Variance and R^2 for Model") |> 
  kable_styling(full_width = F, 
                position = "center", 
                bootstrap_options = c("hover", "condensed")) |> 
  row_spec(0, bold = T, color = "black") |> 
  add_header_above(c(" " = 4))
```

To get the results in the table we had to first find three values from the model. Those were the total variance, the variance from the explanatory value, and the variance from the residual. To get them we used the augment() function from the broom package and then got the sum of the variance in the income, .fitted, and .residual variables. Those are the first three values in the table. The final value, the R\^2, was gotten by dividing the fitted variance by the total variance. This tells us how much of the the variance in the results can be explained by the model. In the context of the data this shows that in 2017 about 43.6% of the differences in mean daily household income between countries can be explained by differences in educational attainment as measured by the proportion of the population with the equivalent of bachelor's degree.

Sources cited:

Pew Research: Source: [https://www.pewresearch.org/social-trends/2024/05/23/public-views-on-the-value-of-a-college-degree/#:\~:text=The%20importance%20of%20a%20four%2Dyear%20college%20degree,-Four%2Din%2Dten&text=About%20half%20of%20Americans%20(49,importance%20hasn't%20really%20changed.](https://www.pewresearch.org/social-trends/2024/05/23/public-views-on-the-value-of-a-college-degree/#:~:text=The%20importance%20of%20a%20four%2Dyear%20college%20degree,-Four%2Din%2Dten&text=About%20half%20of%20Americans%20(49,importance%20hasn't%20really%20changed.))

## Simulation

```{r}
set.seed(151)

prediction = predict(bachelors_lm)
est_sigma = sigma(bachelors_lm)

rand_error = function(x, mean = 0, sd){
  x + rnorm(length(x), mean, sd)
}

bach_sim_resp = tibble(sim_income = rand_error(x = prediction, 
                                               sd = est_sigma))

full_bachelors = country_averages |> 
  bind_cols(bach_sim_resp)
```

```{r}
sim_plot <-full_bachelors |> 
  ggplot(mapping = aes(x = sim_income, y = income))+
  geom_point(alpha = 0.6, size = 2)+
  geom_smooth(method = "lm", color = "blue", se = FALSE)+
  labs(x = "Simulated Income",
       y = "Observed Income",
       title = "Simulated vs Observed Income")+
  theme_minimal()

original_plot <- country_averages |> 
  ggplot(mapping = aes(x = bachelors, y = income))+
  geom_point(alpha = 0.6, size = 2)+
  geom_smooth(method = "lm", color = "blue", se = FALSE)+
  labs(x = "Observed Bachelors vs Observed Income",
       y = "Observed Income",
       title = "observed")+
  theme_minimal()

sim_plot + original_plot
```

```{r}
set.seed(256)
sims = map_dfc(.x = 1:1000, 
               .f = ~ tibble(sim = rand_error(x = prediction, 
                                               sd = est_sigma)))

colnames(sims) <- colnames(sims) |> 
  str_replace(pattern = "\\.\\.\\.",
                  replace = "_")

sims = country_averages |> 
  select(income) |> 
  bind_cols(sims)

sim_r_sq = sims |> 
  map(~ lm(income ~ .x, data = sims)) |> 
  map(glance) |> 
  map_dbl(~ .x$r.squared)

sim_r_sq = sim_r_sq[names(sim_r_sq) != "income"]

tibble(sims = sim_r_sq) |> 
  ggplot(aes(x = sims)) + 
  geom_histogram(fill = "lightblue", 
                 color = "darkgray", 
                 bins = 40) + 
  labs(x = "R² Values", y = "Frequency", 
       title = "Distribution of R² Values from Predictive Checks") +
  theme_minimal()

```






