---
title: "Project Proposal"
author: "Emily Stokes, Sophia Nicotera, Ben Cox, Wendy Tan"
format: revealjs
editor: visual
---

## Project Proposal

In our project, we decided to look at the statistical relationship between education and income across all countries. The rising climate in the United States around education seems to be converging towards a negative attitude, with 49% of Americans thinking that it is less important to have a college degree now than it was in the past (Pew Research). We have selected three datasets: a dataset that captures daily income in a price adjusted form, and two datasets showing the attainment of education: high-school equivalent or bachelor's equivalent.

```{r}
library(readr)
library(tidyverse)
library(kableExtra)
```

We should load in the data now!

```{r}
daily_income <- read_csv("mincpcap_cppp.csv")
high_school_equiv <- read_csv("se_sec_cuat_up_zs.csv")
ba_equiv <- read_csv("se_ter_cuat_ba_zs.csv")
```

Sources cited:

Pew Research: Source: [https://www.pewresearch.org/social-trends/2024/05/23/public-views-on-the-value-of-a-college-degree/#:\~:text=The%20importance%20of%20a%20four%2Dyear%20college%20degree,-Four%2Din%2Dten&text=About%20half%20of%20Americans%20(49,importance%20hasn't%20really%20changed.](https://www.pewresearch.org/social-trends/2024/05/23/public-views-on-the-value-of-a-college-degree/#:~:text=The%20importance%20of%20a%20four%2Dyear%20college%20degree,-Four%2Din%2Dten&text=About%20half%20of%20Americans%20(49,importance%20hasn't%20really%20changed.))
